### 2019-12-15 ic-journal/coarsest.tex
text

The formalisation of this theorem presents some 
delicate aspects. For instance, 
\hl{within} our CCS syntax \hl{which supports} only binary sums,
one way to prove Proposition~\ref{prop:coarsest} is
to add \hl{an additional} hypothesis \hl{that}
 the involved processes \hl{do not use up} all available labels.
 Indeed, this is the standard argument by Milner~\citep[p.~153]{Mil89}.
%
 Formalising this result, however, requires a detailed treatment of
 free and bound names \hl{(of labels)} of CCS
processes, with the restriction operator acting as a binder.
In our actual formalisation, instead,
we assume \hl{a} weaker hypothesis that all \emph{immediate weak} derivatives of
$p$ and $q$ do not \hl{use up} all available labels.
We \hl{may call it} the \emph{free action} property:
\begin{alltt}
   \HOLConst{free_action} \HOLFreeVar{p} \HOLTokenDefEquality{} \HOLSymConst{\HOLTokenExists{}}\HOLBoundVar{a}. \HOLSymConst{\HOLTokenForall{}}\ensuremath{\HOLBoundVar{p}\sp{\prime}}. \HOLSymConst{\HOLTokenNeg{}}\ensuremath{(}\HOLFreeVar{p} \HOLTokenWeakTransBegin\HOLConst{label} \HOLBoundVar{a}\HOLTokenWeakTransEnd \ensuremath{\HOLBoundVar{p}\sp{\prime}}\ensuremath{)}\hfill{[free_action_def]}
\end{alltt}

@@

The formal proofs of Theorem~\ref{thm:coarsestR} and
\ref{thm:coarsestPre} precisely follow Milner~\citep[p.~153--154]{Mil89}.
Although Milner requires a stronger hypothesis: $\mathrm{fn}(p) \cup
\mathrm{fn}(q) \neq \mathscr{L}$ (here $\mathrm{fn}$ stands for \emph{free
  names}), the actual proof essentially requires only the above
free action property.
\hl{Indeed, in the proof one only looks at the immediate weak
derivatives of $p$ and $q$, and only requires that there is an input
or output label that never occurs as a label of the involved weak transitions.
Furthermore, an label $a$ and its complementary $\outC a$ are
considered different: even the involved processes have used up all
available names, including, say, $a$, the whole proof can still complete as
long as $\outC a$ never occurs in the immediate weak
derivatives.}

@@

\hl{Since the type} ``\HOLinline{\ensuremath{(}\ensuremath{\alpha}, \ensuremath{\beta}\ensuremath{)} \HOLTyOp{CCS}}'' is \hl{parameterized with two
type variables, if the type of all label
names $\beta$ has a relatively small cardinality (or just a singleton in the
extreme case), it is quite possible that the involved processes in}
Proposition~\ref{prop:coarsest} used up all available labels, and thus the free
action hypothesis does not hold. In this case, it is still possible to prove
Proposition~\ref{prop:coarsest}, \hl{but due to some
limitations of HOL itself we have to assume that the involved
processes are \emph{finite-state}, i.e. the set of all derivatives is
finite.} The original proof, due to van
Glabbeek~\cite{van2005characterisation}, does not require finite-state CCS.
Here is the main theorem:
\begin{theorem}[\texttt{COARSEST_CONGR_FINITE}]
    \label{thm:coarsestfiniteState}
    For finite-state CCS, $\rapprox$ is the coarsest congruence contained in $\wb$:
\begin{alltt}
\HOLTokenTurnstile{} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{p} \HOLBoundVar{q}. \HOLConst{finite_state} \HOLBoundVar{p} \HOLSymConst{\HOLTokenConj{}} \HOLConst{finite_state} \HOLBoundVar{q} \HOLSymConst{\HOLTokenImp{}} \ensuremath{(}\HOLBoundVar{p} \HOLSymConst{\HOLTokenObsCongr} \HOLBoundVar{q} \HOLSymConst{\HOLTokenEquiv{}} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{r}. \HOLBoundVar{p} \HOLSymConst{\ensuremath{+}} \HOLBoundVar{r} \HOLSymConst{\HOLTokenWeakEQ} \HOLBoundVar{q} \HOLSymConst{\ensuremath{+}} \HOLBoundVar{r}\ensuremath{)}
\end{alltt}
\end{theorem}

The precise definition of \texttt{finite_state} used in above theorem
will be given later. We start with a core lemma (\texttt{PROP3_COMMON}) saying that, for
any two processes $p$ and $q$, if there exists a \emph{stable}
(i.e.~\hl{all immediate derivatives} are $\tau$-free)
 process $k$ which is not bisimilar with any \hl{weak} derivative of $p$ and
 $q$, then \texttt{SUM_EQUIV} indeed implies rooted bisimilarity
 ($\rapprox$) (see~\cite{van2005characterisation,Tian:2017wrba} for
 its proof, both informal and formal versions):
\begin{alltt}
\HOLTokenTurnstile{} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{p} \HOLBoundVar{q}.
       \ensuremath{(}\HOLSymConst{\HOLTokenExists{}}\HOLBoundVar{k}.
            \HOLConst{STABLE} \HOLBoundVar{k} \HOLSymConst{\HOLTokenConj{}} \ensuremath{(}\HOLSymConst{\HOLTokenForall{}}\ensuremath{\HOLBoundVar{p}\sp{\prime}} \HOLBoundVar{u}. \HOLBoundVar{p} \HOLTokenWeakTransBegin\HOLBoundVar{u}\HOLTokenWeakTransEnd \ensuremath{\HOLBoundVar{p}\sp{\prime}} \HOLSymConst{\HOLTokenImp{}} \HOLSymConst{\HOLTokenNeg{}}\ensuremath{(}\ensuremath{\HOLBoundVar{p}\sp{\prime}} \HOLSymConst{\HOLTokenWeakEQ} \HOLBoundVar{k}\ensuremath{)}\ensuremath{)} \HOLSymConst{\HOLTokenConj{}}
            \HOLSymConst{\HOLTokenForall{}}\ensuremath{\HOLBoundVar{q}\sp{\prime}} \HOLBoundVar{u}. \HOLBoundVar{q} \HOLTokenWeakTransBegin\HOLBoundVar{u}\HOLTokenWeakTransEnd \ensuremath{\HOLBoundVar{q}\sp{\prime}} \HOLSymConst{\HOLTokenImp{}} \HOLSymConst{\HOLTokenNeg{}}\ensuremath{(}\ensuremath{\HOLBoundVar{q}\sp{\prime}} \HOLSymConst{\HOLTokenWeakEQ} \HOLBoundVar{k}\ensuremath{)}\ensuremath{)} \HOLSymConst{\HOLTokenImp{}}
       \ensuremath{(}\HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{r}. \HOLBoundVar{p} \HOLSymConst{\ensuremath{+}} \HOLBoundVar{r} \HOLSymConst{\HOLTokenWeakEQ} \HOLBoundVar{q} \HOLSymConst{\ensuremath{+}} \HOLBoundVar{r}\ensuremath{)} \HOLSymConst{\HOLTokenImp{}}
       \HOLBoundVar{p} \HOLSymConst{\HOLTokenObsCongr} \HOLBoundVar{q}\hfill{[PROP3_COMMON]}
\end{alltt}
\begin{alltt}
   \HOLConst{STABLE} \HOLFreeVar{p} \HOLTokenDefEquality{} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{u} \ensuremath{\HOLBoundVar{p}\sp{\prime}}. \HOLFreeVar{p} \HOLTokenTransBegin\HOLBoundVar{u}\HOLTokenTransEnd \ensuremath{\HOLBoundVar{p}\sp{\prime}} \HOLSymConst{\HOLTokenImp{}} \HOLBoundVar{u} \HOLSymConst{\HOLTokenNotEqual{}} \HOLSymConst{\ensuremath{\tau}}\hfill{[STABLE]}
\end{alltt}

To prove Theorem~\ref{thm:coarsestfiniteState}, \hl{it only remains to construct this stable
process $k$ for any two finite-state processes $p$ and $q$.}
\hl{For arbitrary CCS processes, this construction} relies on
arbitrary infinite sums of processes (not within our CCS syntax) and
transfinite induction to obtain
an arbitrary large sequence of processes that are all pairwise
non-bisimilar,
which was firstly introduced by Jan
Willem Klop~(see \cite{van2005characterisation} for some historical notes).
We have only partially formalised
van Glabbeek's proof, mostly because our CCS syntax does not allow infinite
summation (and it is not easy to extend it with this support).
Another more important reason is that, the typed logic
implemented in various HOL systems (including Isabelle/HOL) is not
strong enough to define a type for all possible
ordinals~\cite{norrish2013ordinals} which is required in van
Glabbeek's proof. As the consequence, the formalisation
(Theorem~\ref{thm:coarsestfiniteState})
can only apply to finite-state CCS.

The above core lemma (\texttt{PROP3_COMMON}) requires the
existence of a special CCS process, which is not weakly bisimilar to
any \hl{subprocess emanated} from the two root processes by weak
transitions.
There could be infinitely many such subprocesses, even on finitely
branching processes.
We can, however, consider the equivalence classes of CCS processes
modulo weak bisimilarity.
If there are infinitely many such classes, 
then it will be 
possible to choose one that is distinct from all the (finitely many) states in the
\hl{transition} graphs of the two given processes.
This can be done by following Klop's contruction.
 We call the processes in this construction the ``Klop processes'':
\begin{definition}[Klop processes]
For each ordinal $\lambda$, and an arbitrary chosen action $a \neq \tau$,
define a CCS process $k_\lambda$ as follows:
\begin{itemize}
\item $k_0 = 0$,
\item $k_{\lambda+1} = k_\lambda + a.k_\lambda$ and
\item for $\lambda$ a limit ordinal, $k_\lambda = \sum_{\mu < \lambda}
  k_\mu$ (meaning that $k_\lambda$ is constructed from all graphs
  $k_\mu$ for $\mu < \lambda$ by identifying their root).
\end{itemize}
\end{definition}
When processes are finite-state, that is,
the number of  states in which a process may evolve by performing
transitions is finite, 
we can use  the following subset of Klop processes, 
defined as a recursive function (on natural numbers) in HOL4:
\begin{definition}{(Klop processes as recursive function on natural numbers)}
\begin{alltt}
   \HOLConst{KLOP} \HOLFreeVar{a} \HOLNumLit{0} \HOLTokenDefEquality{} \HOLConst{\ensuremath{\mathbf{0}}}
   \HOLConst{KLOP} \HOLFreeVar{a} \ensuremath{(}\HOLConst{SUC} \HOLFreeVar{n}\ensuremath{)} \HOLTokenDefEquality{} \HOLConst{KLOP} \HOLFreeVar{a} \HOLFreeVar{n} \HOLSymConst{\ensuremath{+}} \HOLConst{label} \HOLFreeVar{a}\HOLSymConst{\ensuremath{\ldotp}}\HOLConst{KLOP} \HOLFreeVar{a} \HOLFreeVar{n}\hfill{[KLOP_def]}
\end{alltt}
\end{definition}

@@

For any ``\HOLinline{\HOLConst{label}\;\HOLFreeVar{a}}'', having a function ``\HOLinline{\HOLConst{KLOP}\;\HOLFreeVar{a}}'' (of the type ``\HOLinline{\HOLTyOp{num} \HOLTokenTransEnd \ensuremath{(}\ensuremath{\alpha}, \ensuremath{\beta}\ensuremath{)} \HOLTyOp{CCS}}'')
\hl{defined on all natural numbers, we can get a countable set of processes
containing all Klop processes built by the same label.
Since the number of all Klop processes in this set is (countably) infinite, and
since they are all pairwise non-bisimiar,
we can always
choose a number that is mapped to a Klop process
that is non-bisimilar with} any derivative of two given 
 finite-state CCS processes $p$ and $q$
(by definition, the number of such derivatives is finite), \emph{even}
if $a$ is the only element of type $\beta$ (i.e. the only label name
in $\mathscr{L}$).
This property is captured by appealing to the following set-theoreric
lamma (see~\cite{Tian:2017wrba} for its proof):
\begin{lemma}
Given an equivalence relation $R$ defined on a type, and two sets $A, B$
of elements in this type, 
if $A$ is finite, $B$ is infinite, and all elements
in $B$ belong to distinct equivalence classes, then there exists an element $k$ in $B$
which is not equivalent to any element in $A$:
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{equivalence} \HOLFreeVar{R} \HOLSymConst{\HOLTokenImp{}}
   \HOLConst{FINITE} \HOLFreeVar{A} \HOLSymConst{\HOLTokenConj{}} \HOLConst{INFINITE} \HOLFreeVar{B} \HOLSymConst{\HOLTokenConj{}} \ensuremath{(}\HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{x} \HOLBoundVar{y}. \HOLBoundVar{x} \HOLSymConst{\HOLTokenIn{}} \HOLFreeVar{B} \HOLSymConst{\HOLTokenConj{}} \HOLBoundVar{y} \HOLSymConst{\HOLTokenIn{}} \HOLFreeVar{B} \HOLSymConst{\HOLTokenConj{}} \HOLBoundVar{x} \HOLSymConst{\HOLTokenNotEqual{}} \HOLBoundVar{y} \HOLSymConst{\HOLTokenImp{}} \HOLSymConst{\HOLTokenNeg{}}\HOLFreeVar{R} \HOLBoundVar{x} \HOLBoundVar{y}\ensuremath{)} \HOLSymConst{\HOLTokenImp{}}
   \HOLSymConst{\HOLTokenExists{}}\HOLBoundVar{k}. \HOLBoundVar{k} \HOLSymConst{\HOLTokenIn{}} \HOLFreeVar{B} \HOLSymConst{\HOLTokenConj{}} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{n}. \HOLBoundVar{n} \HOLSymConst{\HOLTokenIn{}} \HOLFreeVar{A} \HOLSymConst{\HOLTokenImp{}} \HOLSymConst{\HOLTokenNeg{}}\HOLFreeVar{R} \HOLBoundVar{n} \HOLBoundVar{k}\hfill[INFINITE_EXISTS_LEMMA]
\end{alltt}
\end{lemma}
% \begin{proof}
%   We built an explicit mapping $f$ from $A$ to $B$\footnote{There're
%     multiple ways to prove this lemma, a simpler proof is to make a
%     reverse mapping from $B$ to the power set of $A$ (or further use
%     the Axiom of Choice (AC) to make a mapping from $B$ to $A$), then
%     the non-injectivity of this mapping will contradict the fact that
%     all elements in the infinite set are distinct. Our proof doesn't
%     need AC, and it relies on very simple truths about sets.}, for all
%   $x \in A$, $y = f(x)$ if $y \in B$ and $y$ is equivalent with
%   $x$. But it's possible that no element in $B$ is equivalent with
%   $x$, and in this case we just choose an arbitrary element as
%   $f(x)$. Such a mapping is to make sure the range of $f$ always fall
%   into $B$.

@@

To reason about finite-state CCS, we also need to define the concept
of ``finite-state CCS'' as a predicate on CCS processes:
\begin{definition}[finite-state CCS]
  \hl{Let $p$ be an arbitrary CCS process:}
\begin{enumerate}
\item A binary relation \texttt{Reach} is the RTC (reflexive and transitive
  closure) of a relation indicating the existence of a transition between two processes:
\begin{alltt}
\HOLConst{Reach} \HOLTokenDefEquality{} \ensuremath{(}\HOLTokenLambda{}\HOLBoundVar{E} \ensuremath{\HOLBoundVar{E}\sp{\prime}}. \HOLSymConst{\HOLTokenExists{}}\HOLBoundVar{u}. \HOLBoundVar{E} \HOLTokenTransBegin\HOLBoundVar{u}\HOLTokenTransEnd \ensuremath{\HOLBoundVar{E}\sp{\prime}}\ensuremath{)}\HOLSymConst{\HOLTokenSupStar{}}\hfill[Reach_def]
\end{alltt}
\item The set of all derivatives (\texttt{NODES}) of a process is the
  set of all processes reachable from it:
\begin{alltt}
\HOLConst{NODES} \HOLFreeVar{p} \HOLTokenDefEquality{} \HOLTokenLeftbrace{}\HOLBoundVar{q} \HOLTokenBar{} \HOLConst{Reach} \HOLFreeVar{p} \HOLBoundVar{q}\HOLTokenRightbrace{}\hfill[NODES_def]
\end{alltt}
\item A process is \texttt{finite-state} if the set of all derivatives is finite:
\begin{alltt}
\HOLConst{finite_state} \HOLFreeVar{p} \HOLTokenDefEquality{} \HOLConst{FINITE} \ensuremath{(}\HOLConst{NODES} \HOLFreeVar{p}\ensuremath{)}\hfill[finite_state_def]
\end{alltt}
\end{enumerate}
\end{definition}
We rely on various 
 properties of the above definitions, such as the following one:
\begin{proposition}
If $p$ has a weak transition to $q$, then $q$ must be \hl{one of the derivatives} of $p$:
\begin{alltt}
\HOLTokenTurnstile{} \HOLFreeVar{p} \HOLTokenWeakTransBegin\HOLFreeVar{u}\HOLTokenWeakTransEnd \HOLFreeVar{q} \HOLSymConst{\HOLTokenImp{}} \HOLFreeVar{q} \HOLSymConst{\HOLTokenIn{}} \HOLConst{NODES} \HOLFreeVar{p}\hfill[WEAK_TRANS_IN_NODES]
\end{alltt}
\end{proposition}

>>>>

The formalisation of this theorem presents some 
delicate aspects. For instance, 
\hl{within} our CCS syntax \hl{which supports} only binary sums,
one way to prove Proposition~\ref{prop:coarsest} is
to add \hlD{an} hypothesis \hl{that}
 the involved processes \hlD{do not use} all available labels.
 Indeed, this is the standard argument by Milner~\citep[p.~153]{Mil89}.
%
 Formalising this result, however, requires a detailed treatment of
 free and bound names \hl{(of labels)} of CCS
processes, with the restriction operator acting as a binder.
In our actual formalisation, instead,
we assume \hlD{the} weaker hypothesis that all \emph{immediate weak} derivatives of
$p$ and $q$ do not \hlD{use} all available labels.
We \hlD{call this} the \emph{free action} property:
\begin{alltt}
   \HOLConst{free_action} \HOLFreeVar{p} \HOLTokenDefEquality{} \HOLSymConst{\HOLTokenExists{}}\HOLBoundVar{a}. \HOLSymConst{\HOLTokenForall{}}\ensuremath{\HOLBoundVar{p}\sp{\prime}}. \HOLSymConst{\HOLTokenNeg{}}\ensuremath{(}\HOLFreeVar{p} \HOLTokenWeakTransBegin\HOLConst{label} \HOLBoundVar{a}\HOLTokenWeakTransEnd \ensuremath{\HOLBoundVar{p}\sp{\prime}}\ensuremath{)}\hfill{[free_action_def]}
\end{alltt}

@@

The formal proofs of Theorem~\ref{thm:coarsestR} and
\ref{thm:coarsestPre} precisely follow Milner~\citep[p.~153--154]{Mil89}.
Although Milner requires a stronger hypothesis: $\mathrm{fn}(p) \cup
\mathrm{fn}(q) \neq \mathscr{L}$ (here $\mathrm{fn}$ stands for \emph{free
  names}), the actual proof essentially requires only the above
free action property.
\hl{Indeed, in the proof one only looks at the immediate weak
derivatives of $p$ and $q$, and only requires that there is an input
or output label that never occurs as a label of the involved transitions.
% Furthermore, an label $a$ and its complementary $\outC a$ are
% considered different: even the involved processes have used up all
% available names, including, say, $a$, the whole proof can still complete as
% long as $\outC a$ never occurs in the immediate weak
% derivatives.
}

@@

\hlD{As the type} ``\HOLinline{\ensuremath{(}\ensuremath{\alpha}, \ensuremath{\beta}\ensuremath{)} \HOLTyOp{CCS}}'' is \hl{parameterized on two
type variables, if the type of all label
names $\beta$ has a  small cardinality (a singleton in the
extreme case), it is  possible that the processes of}
Proposition~\ref{prop:coarsest} use all available labels, and thus the free
action hypothesis does not hold. In this case, it is still possible to prove
Proposition~\ref{prop:coarsest};  \hlD{however,  due to some
limitations of HOL itself we have to assume that the
processes are \emph{finite-state}, i.e. the set of all their derivatives is
finite.} The original proof, due to van
Glabbeek~\cite{van2005characterisation}, does not require finite-state CCS.
Here is the main theorem:
\begin{theorem}[\texttt{COARSEST_CONGR_FINITE}]
    \label{thm:coarsestfiniteState}
    For finite-state CCS, $\rapprox$ is the coarsest congruence contained in $\wb$:
\begin{alltt}
\HOLTokenTurnstile{} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{p} \HOLBoundVar{q}. \HOLConst{finite_state} \HOLBoundVar{p} \HOLSymConst{\HOLTokenConj{}} \HOLConst{finite_state} \HOLBoundVar{q} \HOLSymConst{\HOLTokenImp{}} \ensuremath{(}\HOLBoundVar{p} \HOLSymConst{\HOLTokenObsCongr} \HOLBoundVar{q} \HOLSymConst{\HOLTokenEquiv{}} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{r}. \HOLBoundVar{p} \HOLSymConst{\ensuremath{+}} \HOLBoundVar{r} \HOLSymConst{\HOLTokenWeakEQ} \HOLBoundVar{q} \HOLSymConst{\ensuremath{+}} \HOLBoundVar{r}\ensuremath{)}
\end{alltt}
\end{theorem}

The precise definition of \texttt{finite_state} used in above theorem
will be given later. We start with a core lemma (\texttt{PROP3_COMMON}) saying that, for
any two processes $p$ and $q$, if there exists a \emph{stable}
(i.e.~\hl{without  $\tau$ transitions})
 process which is not bisimilar with any \hl{weak} derivative of $p$ and
 $q$, then \HOLinline{\HOLConst{SUM_EQUIV}} indeed implies rooted bisimilarity
 ($\rapprox$)~\cite{van2005characterisation,Tian:2017wrba}:
\begin{alltt}
\HOLTokenTurnstile{} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{p} \HOLBoundVar{q}.
       \ensuremath{(}\HOLSymConst{\HOLTokenExists{}}\HOLBoundVar{k}.
            \HOLConst{STABLE} \HOLBoundVar{k} \HOLSymConst{\HOLTokenConj{}} \ensuremath{(}\HOLSymConst{\HOLTokenForall{}}\ensuremath{\HOLBoundVar{p}\sp{\prime}} \HOLBoundVar{u}. \HOLBoundVar{p} \HOLTokenWeakTransBegin\HOLBoundVar{u}\HOLTokenWeakTransEnd \ensuremath{\HOLBoundVar{p}\sp{\prime}} \HOLSymConst{\HOLTokenImp{}} \HOLSymConst{\HOLTokenNeg{}}\ensuremath{(}\ensuremath{\HOLBoundVar{p}\sp{\prime}} \HOLSymConst{\HOLTokenWeakEQ} \HOLBoundVar{k}\ensuremath{)}\ensuremath{)} \HOLSymConst{\HOLTokenConj{}}
            \HOLSymConst{\HOLTokenForall{}}\ensuremath{\HOLBoundVar{q}\sp{\prime}} \HOLBoundVar{u}. \HOLBoundVar{q} \HOLTokenWeakTransBegin\HOLBoundVar{u}\HOLTokenWeakTransEnd \ensuremath{\HOLBoundVar{q}\sp{\prime}} \HOLSymConst{\HOLTokenImp{}} \HOLSymConst{\HOLTokenNeg{}}\ensuremath{(}\ensuremath{\HOLBoundVar{q}\sp{\prime}} \HOLSymConst{\HOLTokenWeakEQ} \HOLBoundVar{k}\ensuremath{)}\ensuremath{)} \HOLSymConst{\HOLTokenImp{}}
       \ensuremath{(}\HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{r}. \HOLBoundVar{p} \HOLSymConst{\ensuremath{+}} \HOLBoundVar{r} \HOLSymConst{\HOLTokenWeakEQ} \HOLBoundVar{q} \HOLSymConst{\ensuremath{+}} \HOLBoundVar{r}\ensuremath{)} \HOLSymConst{\HOLTokenImp{}}
       \HOLBoundVar{p} \HOLSymConst{\HOLTokenObsCongr} \HOLBoundVar{q}\hfill{[PROP3_COMMON]}
\end{alltt}
\begin{alltt}
   \HOLConst{STABLE} \HOLFreeVar{p} \HOLTokenDefEquality{} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{u} \ensuremath{\HOLBoundVar{p}\sp{\prime}}. \HOLFreeVar{p} \HOLTokenTransBegin\HOLBoundVar{u}\HOLTokenTransEnd \ensuremath{\HOLBoundVar{p}\sp{\prime}} \HOLSymConst{\HOLTokenImp{}} \HOLBoundVar{u} \HOLSymConst{\HOLTokenNotEqual{}} \HOLSymConst{\ensuremath{\tau}}\hfill{[STABLE]}
\end{alltt}

To prove Theorem~\ref{thm:coarsestfiniteState}, \hl{it only remains to construct such stable
process $k$ for any two finite-state processes $p$ and $q$.}
\hl{For arbitrary CCS processes, this construction} relies on
arbitrary infinite sums of processes (not within our CCS syntax) and
transfinite induction to obtain
an arbitrary large sequence of processes that are all pairwise
non-bisimilar,
which was firstly introduced by Jan
Willem Klop~(see \cite{van2005characterisation} for some historical notes).
We have only partially formalised
van Glabbeek's proof, mostly because our CCS syntax does not allow infinite
summation (and it is not easy to extend it with this support).
Another more important reason is that, the typed logic
implemented in various HOL systems (including Isabelle/HOL) is not
strong enough to define a type for all possible
ordinals~\cite{norrish2013ordinals} which is required in van
Glabbeek's proof. As the consequence, the formalisation
(Theorem~\ref{thm:coarsestfiniteState})
can only apply to finite-state CCS.

The above core lemma (\texttt{PROP3_COMMON}) requires the
existence of a special CCS process, which is not weakly bisimilar to
any \hlD{weak derivative} of the two root processes.
There could be infinitely many such subprocesses, even on finitely
branching processes.
We can, however, consider the equivalence classes of CCS processes
modulo weak bisimilarity.
If there are infinitely many such classes, 
then it will be 
possible to choose one that is distinct from all the (finitely many) states in the
\hl{transition} graphs of the two given processes.
This can be done by following Klop's contruction.
 We call the processes in this construction the ``Klop processes'':
\begin{definition}[Klop processes]
For each ordinal $\lambda$, and an arbitrary chosen action $a \neq \tau$,
define a CCS process $k_\lambda$ as follows:
\begin{itemize}
\item $k_0 = 0$,
\item $k_{\lambda+1} = k_\lambda + a.k_\lambda$ and
\item for $\lambda$ a limit ordinal, $k_\lambda = \sum_{\mu < \lambda}
  k_\mu$ (meaning that $k_\lambda$ is constructed from all graphs
  $k_\mu$ for $\mu < \lambda$ by identifying their root).
\end{itemize}
\end{definition}
When processes are finite-state, that is,
the number of  states in which a process may evolve by performing
transitions is finite, 
we can use  the following subset of Klop processes, 
defined as a recursive function (on natural numbers) in HOL4:
\begin{definition}{(Klop processes as recursive function on natural numbers)}
\begin{alltt}
   \HOLConst{KLOP} \HOLFreeVar{a} \HOLNumLit{0} \HOLTokenDefEquality{} \HOLConst{\ensuremath{\mathbf{0}}}
   \HOLConst{KLOP} \HOLFreeVar{a} \ensuremath{(}\HOLConst{SUC} \HOLFreeVar{n}\ensuremath{)} \HOLTokenDefEquality{} \HOLConst{KLOP} \HOLFreeVar{a} \HOLFreeVar{n} \HOLSymConst{\ensuremath{+}} \HOLConst{label} \HOLFreeVar{a}\HOLSymConst{\ensuremath{\ldotp}}\HOLConst{KLOP} \HOLFreeVar{a} \HOLFreeVar{n}\hfill{[KLOP_def]}
\end{alltt}
\end{definition}

@@

For any ``\HOLinline{\HOLConst{label}\;\HOLFreeVar{a}}'', having a function ``\HOLinline{\HOLConst{KLOP}\;\HOLFreeVar{a}}'' (of  type ``\HOLinline{\HOLTyOp{num} \HOLTokenTransEnd \ensuremath{(}\ensuremath{\alpha}, \ensuremath{\beta}\ensuremath{)} \HOLTyOp{CCS}}'')
\hlD{defined on the natural numbers, we obtain a countable set of processes
with all Klop processes built from the same label.
Since the number of all Klop processes in this set is (countably) infinite, and
since they are all pairwise non-bisimiar,
we can always
choose a number that is mapped to a Klop process
that is non-bisimilar with} any derivative of two given 
 (finite-state) processes $p$ and $q$,
%(by definition, the number of such derivatives is finite), 
({even}
when $a$ is the only element of type $\beta$, i.e. the only label name
in $\mathscr{L}$).
This property is captured by appealing to the following set-theoreric
lamma (see~\cite{Tian:2017wrba} for its proof):
\begin{lemma}
Given an equivalence relation $R$ defined on a type, and two sets $A, B$
of elements in this type, 
if $A$ is finite, $B$ is infinite, and all elements
in $B$ belong to distinct equivalence classes, then there exists an element $k$ in $B$
which is not equivalent to any element in $A$:
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{equivalence} \HOLFreeVar{R} \HOLSymConst{\HOLTokenImp{}}
   \HOLConst{FINITE} \HOLFreeVar{A} \HOLSymConst{\HOLTokenConj{}} \HOLConst{INFINITE} \HOLFreeVar{B} \HOLSymConst{\HOLTokenConj{}} \ensuremath{(}\HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{x} \HOLBoundVar{y}. \HOLBoundVar{x} \HOLSymConst{\HOLTokenIn{}} \HOLFreeVar{B} \HOLSymConst{\HOLTokenConj{}} \HOLBoundVar{y} \HOLSymConst{\HOLTokenIn{}} \HOLFreeVar{B} \HOLSymConst{\HOLTokenConj{}} \HOLBoundVar{x} \HOLSymConst{\HOLTokenNotEqual{}} \HOLBoundVar{y} \HOLSymConst{\HOLTokenImp{}} \HOLSymConst{\HOLTokenNeg{}}\HOLFreeVar{R} \HOLBoundVar{x} \HOLBoundVar{y}\ensuremath{)} \HOLSymConst{\HOLTokenImp{}}
   \HOLSymConst{\HOLTokenExists{}}\HOLBoundVar{k}. \HOLBoundVar{k} \HOLSymConst{\HOLTokenIn{}} \HOLFreeVar{B} \HOLSymConst{\HOLTokenConj{}} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{n}. \HOLBoundVar{n} \HOLSymConst{\HOLTokenIn{}} \HOLFreeVar{A} \HOLSymConst{\HOLTokenImp{}} \HOLSymConst{\HOLTokenNeg{}}\HOLFreeVar{R} \HOLBoundVar{n} \HOLBoundVar{k}\hfill[INFINITE_EXISTS_LEMMA]
\end{alltt}
\end{lemma}
% \begin{proof}
%   We built an explicit mapping $f$ from $A$ to $B$\footnote{There're
%     multiple ways to prove this lemma, a simpler proof is to make a
%     reverse mapping from $B$ to the power set of $A$ (or further use
%     the Axiom of Choice (AC) to make a mapping from $B$ to $A$), then
%     the non-injectivity of this mapping will contradict the fact that
%     all elements in the infinite set are distinct. Our proof doesn't
%     need AC, and it relies on very simple truths about sets.}, for all
%   $x \in A$, $y = f(x)$ if $y \in B$ and $y$ is equivalent with
%   $x$. But it's possible that no element in $B$ is equivalent with
%   $x$, and in this case we just choose an arbitrary element as
%   $f(x)$. Such a mapping is to make sure the range of $f$ always fall
%   into $B$.

@@

To reason about finite-state CCS, we also need to define the concept
of ``finite-state CCS'' as a predicate on CCS processes:
\begin{definition}[finite-state CCS]
%  \hl{Let $p$ be an arbitrary CCS process:}
\begin{enumerate}
\item A binary relation \texttt{Reach} is the RTC (reflexive and transitive
  closure) of a relation indicating the existence of a transition between two processes:
\begin{alltt}
\HOLConst{Reach} \HOLTokenDefEquality{} \ensuremath{(}\HOLTokenLambda{}\HOLBoundVar{E} \ensuremath{\HOLBoundVar{E}\sp{\prime}}. \HOLSymConst{\HOLTokenExists{}}\HOLBoundVar{u}. \HOLBoundVar{E} \HOLTokenTransBegin\HOLBoundVar{u}\HOLTokenTransEnd \ensuremath{\HOLBoundVar{E}\sp{\prime}}\ensuremath{)}\HOLSymConst{\HOLTokenSupStar{}}\hfill[Reach_def]
\end{alltt}
\item The set of all derivatives (\texttt{NODES}) of a process is the
  set of all processes reachable from it:
\begin{alltt}
\HOLConst{NODES} \HOLFreeVar{p} \HOLTokenDefEquality{} \HOLTokenLeftbrace{}\HOLBoundVar{q} \HOLTokenBar{} \HOLConst{Reach} \HOLFreeVar{p} \HOLBoundVar{q}\HOLTokenRightbrace{}\hfill[NODES_def]
\end{alltt}
\item A process is \texttt{finite-state} if the set of all derivatives is finite:
\begin{alltt}
\HOLConst{finite_state} \HOLFreeVar{p} \HOLTokenDefEquality{} \HOLConst{FINITE} \ensuremath{(}\HOLConst{NODES} \HOLFreeVar{p}\ensuremath{)}\hfill[finite_state_def]
\end{alltt}
\end{enumerate}
\end{definition}
We rely on various 
 properties of the above definitions, such as the following one:
\begin{proposition}
If $p$ has a weak transition to $q$, then $q$ is among \hlD{the derivatives} of $p$:
\begin{alltt}
\HOLTokenTurnstile{} \HOLFreeVar{p} \HOLTokenWeakTransBegin\HOLFreeVar{u}\HOLTokenWeakTransEnd \HOLFreeVar{q} \HOLSymConst{\HOLTokenImp{}} \HOLFreeVar{q} \HOLSymConst{\HOLTokenIn{}} \HOLConst{NODES} \HOLFreeVar{p}\hfill[WEAK_TRANS_IN_NODES]
\end{alltt}
\end{proposition}
